### 2020-03-07 Hands-on Machine Learning/17-20_Unsupervised-Learning.rmd
homl unsupervised
homl unsupervised

---
title: ''
mainfont: Arial
fontsize: 12pt
documentclass: report
header-includes:
- \PassOptionsToPackage{table}{xcolor}
- \usepackage{caption}
- \usepackage{amssymb}
- \usepackage{booktabs}
- \usepackage{longtable}
- \usepackage{array}
- \usepackage{multirow}
- \usepackage{wrapfig}
- \usepackage{float}
- \usepackage{colortbl}
- \usepackage{pdflscape}
- \usepackage{tabu}
- \usepackage{threeparttable}
- \usepackage{threeparttablex}
- \usepackage[normalem]{ulem}
- \usepackage{makecell}
- \usepackage[table]{xcolor}
- \usepackage{fancyhdr}
- \usepackage{boldline}
- \usepackage{tipa}
   \definecolor{headergrey}{HTML}{545454}
   \definecolor{msdblue}{HTML}{1C93D1}
   \pagestyle{fancy}
   \setlength\headheight{30pt}
   \rhead{\color{headergrey}\today}
   \fancyhead[L]{\color{headergrey}Moretz, Brandon}
   \fancyhead[C]{\Large\bfseries\color{headergrey}Interpretable Machine Learning}
   \rfoot{\color{headergrey}\thepage}
   \lfoot{\color{headergrey}Chapter 16}
   \fancyfoot[C]{\rmfamily\color{headergrey}Hands-On Machine Learning}
geometry: left = 1cm, right = 1cm, top = 2cm, bottom = 3cm
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document:
    fig_caption: yes
    latex_engine: xelatex
editor_options: 
  chunk_output_type: console
fig_width: 9
fig_height: 3.5
---

@@

## Interpretable Machine Learning

@@

```{r, ames_data}

ames <- AmesHousing::make_ames()
ames_h2o <- as.h2o(ames)

@@

```{r, data_recipe}

# ensure consistent categorical levels
blueprint <- recipe(Sale_Price ~., data = ames_train) %>%
  step_other(all_nominal(), threshold = 0.005)

# Create training / test h2o frames
train_h2o <- prep(blueprint, training = ames_train, retain = T) %>%
  juice() %>%
  as.h2o()

test_h2o <- prep(blueprint, training = ames_train) %>%
  bake(new_data = ames_test) %>%
  as.h2o()

Y <- "Sale_Price"
X <- setdiff(names(ames_train), Y)
```

### h2o ML setup

```{r, train_h2o_models}

# Train & cross-validate a GLM model
best_glm <- h2o.glm(
  x = X, y = Y, training_frame = train_h2o, alpha = 0.1,
  remove_collinear_columns = TRUE, nfolds = 10, fold_assignment = "Modulo",
  keep_cross_validation_predictions = TRUE, seed = 123
)

# Train & cross-validate a RF model
best_rf <- h2o.randomForest(
  x = X, y = Y, training_frame = train_h2o, ntrees = 1000, mtries = 20,
  max_depth = 30, min_rows = 1, sample_rate = 0.8, nfolds = 10,
  fold_assignment = "Modulo", keep_cross_validation_predictions = TRUE,
  seed = 123, stopping_rounds = 5, stopping_metric = "RMSE",
  stopping_tolerance = 0
)

# Train & cross-validate a GBM model
best_gbm <- h2o.gbm(
  x = X, y = Y, training_frame = train_h2o, ntrees = 5000, learn_rate = 0.01,
  max_depth = 7, min_rows = 5, sample_rate = 0.8, nfolds = 10,
  fold_assignment = "Modulo", keep_cross_validation_predictions = TRUE,
  seed = 123, stopping_rounds = 5, stopping_metric = "RMSE",
  stopping_tolerance = 0
)

# Train & cross-validate an XGBoost model

# not avaliable on windows, yet..(3/2/20)

#best_xgb <- h2o.xgboost(
#  x = X, y = Y, training_frame = train_h2o, ntrees = 5000, learn_rate = 0.05,
#  max_depth = 3, min_rows = 3, sample_rate = 0.8, categorical_encoding = "Enum",
#  nfolds = 10, fold_assignment = "Modulo", 
#  keep_cross_validation_predictions = TRUE, seed = 123, stopping_rounds = 50,
#  stopping_metric = "RMSE", stopping_tolerance = 0
#)

# Train a stacked tree ensemble
ensemble_tree <- h2o.stackedEnsemble(
  x = X, y = Y, training_frame = train_h2o, model_id = "my_tree_ensemble_01",
  base_models = list(best_glm, best_rf, best_gbm),
  metalearner_algorithm = "drf"
)
```

### Local Interpretation

Predictions using ensemble tree:

```{r, local_predictions}

# Compute predictions
predictions <- predict(ensemble_tree, train_h2o) %>% as.vector()

# Print the highest and lowest predicted sales price
paste("Observation", which.max(predictions), 
      "has a predicted sale price of", scales::dollar(max(predictions)))

paste("Observation", which.min(predictions), 
      "has a predicted sale price of", scales::dollar(min(predictions)))

# Grab feature values for observations with min/max predicted sales price
high_ob <- as.data.frame(train_h2o)[which.max(predictions), ] %>% select(-Sale_Price)
low_ob  <- as.data.frame(train_h2o)[which.min(predictions), ] %>% select(-Sale_Price)
```

### Model specific vs. model agnostic

```{r, model_features}

# 1) create a data frame with just the features
features <- as.data.frame(train_h2o) %>% select(-Sale_Price)

# 2) Create a vector with the actual responses
response <- as.data.frame(train_h2o) %>% pull(Sale_Price)

# 3) Create custom predict function that returns the predicted values as a vector
pred <- function(object, newdata)  {
  results <- as.vector(h2o.predict(object, as.h2o(newdata)))
  return(results)
}

# Example of prediction output
pred(ensemble_tree, features) %>% head()
```

```{r, iml_dalex_components}
# iml model agnostic object
components_iml <- Predictor$new(
  model = ensemble_tree, 
  data = features, 
  y = response, 
  predict.fun = pred
)

# DALEX model agnostic object
components_dalex <- DALEX::explain(
  model = ensemble_tree,
  data = features,
  y = response,
  predict_function = pred
)
```

### Permutation based feature importance

Implementation

```{r, vip_plot}

vip(
  ensemble_tree,
  train = as.data.frame(train_h2o),
  method = "permute",
  target = "Sale_Price",
  metric = "RMSE",
  nsim = 5,
  sample_frac = 0.5,
  pred_wrapper = pred
)

@@

Partial dependence

```{r, pdp_partial}

# Compute partial dependence values

pd_values <- partial(
  ensemble_tree,
  train = as.data.frame(train_h2o), 
  pred.var = "Gr_Liv_Area",
  pred.fun = function(object, newdata) {
      results <- mean(as.vector(h2o.predict(object, as.h2o(newdata))))
      return(results)
    },
  grid.resolution = 20
)

head(pd_values)  # take a peak

# Partial dependence plot
autoplot(pd_values, rug = TRUE, train = as.data.frame(train_h2o))
```

### Individual conditional expectation

```{r, ice_curves}

# Construct ICE curves
ice_non_centered <- partial(
  ensemble_tree,
  train = as.data.frame(train_h2o), 
  pred.var = "Gr_Liv_Area",
  pred.fun = pred,
  grid.resolution = 20
) %>%
  autoplot(alpha = 0.05, center = FALSE) +
  ggtitle("A) Non-centered ICE curves")

# Construct c-ICE curves
ice_centered <- partial(
  ensemble_tree,
  train = as.data.frame(train_h2o), 
  pred.var = "Gr_Liv_Area",
  pred.fun = pred,
  grid.resolution = 20
) %>%
  autoplot(alpha = 0.05, center = TRUE) +
  ggtitle("B) Centered ICE curves")

# Display plots side by side
gridExtra::grid.arrange(ice_non_centered, ice_centered, ncol = 2)
```

Implementation

### Feature interactions

```{r, pdp_ice_curves}
# Construct c-ICE curves
partial(
  ensemble_tree,
  train = as.data.frame(train_h2o), 
  pred.var = "Gr_Liv_Area",
  pred.fun = pred,
  grid.resolution = 20,
  plot = TRUE,
  center = TRUE,
  plot.engine = "ggplot2"
)
```

```{r, interactions}

interact <- Interaction$new(components_iml)

interact$results %>%
  arrange(desc(.interaction)) %>%
  head()

plot(interact)
```

### Local interpretable model-agnostic explanations

```{r, lime_explainer}

# Create explainer object
components_lime <- lime(
  x = features,
  model = ensemble_tree, 
  n_bins = 10
)

class(components_lime)

# Use LIME to explain previously defined instances: high_ob and low_ob
lime_explanation <- lime::explain(
  x = rbind(high_ob, low_ob), 
  explainer = components_lime, 
  n_permutations = 10,
  dist_fun = "gower",
  kernel_width = 0.25,
  n_features = 10, 
  feature_select = "highest_weights"
)

plot_features(lime_explanation, ncol = 1)
```

### Tuning

```{r, tuning}
# Tune the LIME algorithm a bit
lime_explanation2 <- explain(
  x = rbind(high_ob, low_ob), 
  explainer = components_lime, 
  n_permutations = 5000,
  dist_fun = "euclidean",
  kernel_width = 0.75,
  n_features = 10, 
  feature_select = "lasso_path"
)

# Plot the results
plot_features(lime_explanation2, ncol = 1)

# Two-way PDP using iml
interaction_pdp <- Partial$new(
  components_iml, 
  c("First_Flr_SF", "Overall_Qual"), 
  ice = FALSE, 
  grid.size = 20
)

labels <- interaction_pdp$results %>% filter(First_Flr_SF == max(First_Flr_SF))
plot(interaction_pdp) + 
  ggrepel::geom_label_repel(
    data = labels, 
    aes(label = Overall_Qual),
    label.size = .05, 
    label.padding = .15
)

# Construct c-ICE curves
partial(
  ensemble_tree,
  train = as.data.frame(train_h2o), 
  pred.var = "Gr_Liv_Area",
  pred.fun = pred,
  grid.resolution = 20,
  plot = TRUE,
  center = TRUE,
  plot.engine = "ggplot2"
)
```

### Shapley values

```{r}
# Compute (approximate) Shapley values
(shapley <- Shapley$new(components_iml, x.interest = high_ob, sample.size = 1000))

# Plot results
plot(shapley)

# Reuse existing object
shapley$explain(x.interest = low_ob)

# Plot results
shapley$results %>%
  top_n(25, wt = abs(phi)) %>%
  ggplot(aes(phi, reorder(feature.value, phi), color = phi > 0)) +
  geom_point(show.legend = FALSE)
```

>>>>

---
title: ''
mainfont: Arial
fontsize: 12pt
documentclass: report
header-includes:
- \PassOptionsToPackage{table}{xcolor}
- \usepackage{caption}
- \usepackage{amssymb}
- \usepackage{booktabs}
- \usepackage{longtable}
- \usepackage{array}
- \usepackage{multirow}
- \usepackage{wrapfig}
- \usepackage{float}
- \usepackage{colortbl}
- \usepackage{pdflscape}
- \usepackage{tabu}
- \usepackage{threeparttable}
- \usepackage{threeparttablex}
- \usepackage[normalem]{ulem}
- \usepackage{makecell}
- \usepackage[table]{xcolor}
- \usepackage{fancyhdr}
- \usepackage{boldline}
- \usepackage{tipa}
   \definecolor{headergrey}{HTML}{545454}
   \definecolor{msdblue}{HTML}{1C93D1}
   \pagestyle{fancy}
   \setlength\headheight{30pt}
   \rhead{\color{headergrey}\today}
   \fancyhead[L]{\color{headergrey}Moretz, Brandon}
   \fancyhead[C]{\Large\bfseries\color{headergrey}Principal Component Analysis}
   \rfoot{\color{headergrey}\thepage}
   \lfoot{\color{headergrey}Chapters 17-20}
   \fancyfoot[C]{\rmfamily\color{headergrey}Hands-On Machine Learning}
geometry: left = 1cm, right = 1cm, top = 2cm, bottom = 3cm
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document:
    fig_caption: yes
    latex_engine: xelatex
editor_options: 
  chunk_output_type: console
fig_width: 9
fig_height: 3.5
---

@@

## Unsupervised Learning

@@

```{r, ames_data}
ames <- AmesHousing::make_ames()
ames_h2o <- as.h2o(ames)

@@

Market Basket:

```{r}
url <- "https://koalaverse.github.io/homlr/data/my_basket.csv"

my_basket <- readr::read_csv(url)

dim(my_basket)
```

MNIST:

```{r}
mnist <- dslabs::read_mnist()
```

### Principal Components Analysis

h2o PCA

```{r}
my_basket_h2o <- as.h2o(my_basket)

# PCA
my_pca <- h2o.prcomp(
  training_frame = my_basket_h2o,
  pca_method = "GramSVD",
  k = ncol(my_basket_h2o),
  transform = "STANDARDIZE",
  impute_missing = T,
  max_runtime_secs = 1000
)
```

Loadings

```{r}
my_pca@model$eigenvectors %>%
  as.data.frame() %>%
  mutate(feature = row.names(.)) %>%
  ggplot(aes(pc1, reorder(feature, pc1))) +
  geom_point()
```

```{r}
my_pca@model$eigenvectors %>%
  as.data.frame() %>%
  mutate(feature = row.names(.)) %>%
  ggplot(aes(pc1, pc2, label = feature)) +
  geom_text()
```

Eigenvalue Criterion

```{r}
eigen <- my_pca@model$importance["Standard deviation", ] %>%
  as.vector() %>%
  .^2

sum(eigen)

which(eigen >= 1)
```

Proportion of Variance Explained

```{r}
data.frame(
  PC = my_pca@model$importance %>% seq_along(),
  PVE = my_pca@model$importance %>% .[2,] %>% unlist(),
  CVE = my_pca@model$importance %>% .[3,] %>% unlist()
) %>%
  tidyr::gather(metric, variance_explained, -PC) %>%
  ggplot(aes(PC, variance_explained)) +
  geom_point() +
  facet_wrap( ~ metric, ncol = 1, scales = "free")
```

```{r}
data.frame(
  PC = my_pca@model$importance %>% seq_along,
  PVE = my_pca@model$importance %>% .[2,] %>% unlist()
) %>%
  ggplot(aes(PC, PVE, group = 1, label = PC)) +
  geom_point() +
  geom_line() +
  geom_text(nudge_y = 0.002)
```

## Generalized Low Rank Models

h2o implementation

```{r}
basic_glrm <- h2o.glrm(
  training_frame = my_basket_h2o,
  k = 20,
  loss = "Quadratic",
  regularization_x = "None",
  regularization_y = "None",
  transform = "STANDARDIZE",
  max_iterations = 2000,
  seed = 123
)

summary(basic_glrm)

plot(basic_glrm)
```

```{r}
data.frame(
  PC = basic_glrm@model$importance %>% seq_along(),
  PVE = basic_glrm@model$importance %>% .[2,] %>% unlist(),
  CVE = basic_glrm@model$importance %>% .[3,] %>% unlist()
) %>%
  gather(metric, variance_explained, -PC) %>%
  ggplot(aes(PC, variance_explained)) +
  geom_point() +
  facet_wrap( ~ metric, ncol = 1, scales = "free")
```

```{r}
t(basic_glrm@model$archetypes)[1:5, 1:5]
```

```{r}
p1 <- t(basic_glrm@model$archetypes) %>%
  as.data.frame() %>%
  mutate(feature = row.names(.)) %>%
  ggplot(aes(Arch1, reorder(feature, Arch1))) +
  geom_point()

p2 <- t(basic_glrm@model$archetypes) %>%
  as.data.frame() %>%
  mutate(feature = row.names(.)) %>%
  ggplot(aes(Arch1, Arch2, label = feature)) +
  geom_text()

gridExtra::grid.arrange(p1, p2, nrow = 1)
```

```{r}
k8_glrm <- h2o.glrm(
  training_frame = my_basket_h2o,
  k = 8,
  loss = "Quadratic",
  regularization_x = "None",
  regularization_y = "None",
  transform = "STANDARDIZE",
  max_iterations = 2000,
  seed = 123
)

my_reconstruction <- h2o.reconstruct(k8_glrm, my_basket_h2o,
                                     reverse_transform = T)

my_reconstruction[1:5, 1:4] %>% round(0)
```

```{r}
# Use non-negative regularization
k8_glrm_regularized <- h2o.glrm(
  training_frame = my_basket_h2o,
  k = 8, 
  loss = "Quadratic",
  regularization_x = "NonNegative", 
  regularization_y = "NonNegative",
  gamma_x = 0.5,
  gamma_y = 0.5,
  transform = "STANDARDIZE", 
  max_iterations = 2000,
  seed = 123
)

# Show predicted values
predict(k8_glrm_regularized, my_basket_h2o)[1:5, 1:5]

# Compare regularized versus non-regularized loss
par(mfrow = c(1, 2))
plot(k8_glrm)
plot(k8_glrm_regularized)
```

```{r}
# Split data into train & validation
split <- h2o.splitFrame(my_basket_h2o, ratios = 0.75, seed = 123)
train <- split[[1]]
valid <- split[[2]]

# Create hyperparameter search grid
params <- expand.grid(
  regularization_x = c("None", "NonNegative", "L1"),
  regularization_y = c("None", "NonNegative", "L1"),
  gamma_x = seq(0, 1, by = .25),
  gamma_y = seq(0, 1, by = .25),
  error = 0,
  stringsAsFactors = FALSE
)

# Perform grid search
for(i in seq_len(nrow(params))) {

  # Create model
  glrm_model <- h2o.glrm(
    training_frame = train,
    k = 8, 
    loss = "Quadratic",
    regularization_x = params$regularization_x[i], 
    regularization_y = params$regularization_y[i],
    gamma_x = params$gamma_x[i],
    gamma_y = params$gamma_y[i],
    transform = "STANDARDIZE", 
    max_runtime_secs = 1000,
    seed = 123
  )

  # Predict on validation set and extract error
  validate <- h2o.performance(glrm_model, valid)
  params$error[i] <- validate@metrics$numerr
}

# Look at the top 10 models with the lowest error rate
params %>%
  arrange(error) %>%
  head(10)
```

```{r}
# Apply final model with optimal hyperparamters
final_glrm_model <- h2o.glrm(
  training_frame = my_basket_h2o,
  k = 8, 
  loss = "Quadratic",
  regularization_x = "L1", 
  regularization_y = "NonNegative",
  gamma_x = 1,
  gamma_y = 0.25,
  transform = "STANDARDIZE", 
  max_iterations = 2000,
  seed = 123
)

# New observations to score
new_observations <- as.h2o(sample_n(my_basket, 2))

# Basic scoring
predict(final_glrm_model, new_observations) %>% round(0)
```

### Autoencoders

```{r}
# Convert mnist features to an h2o input data set
features <- as.h2o(mnist$train$images)

# Train an autoencoder
ae1 <- h2o.deeplearning(
  x = seq_along(features),
  training_frame = features,
  autoencoder = TRUE,
  hidden = 2,
  activation = 'Tanh',
  sparse = TRUE
)

# Extract the deep features
ae1_codings <- h2o.deepfeatures(ae1, features, layer = 1)
ae1_codings
```

### Stacked Autoencoders

```{r}
# Hyperparameter search grid
hyper_grid <- list(hidden = list(
  c(50),
  c(100), 
  c(300, 100, 300),
  c(100, 50, 100),
  c(250, 100, 50, 100, 250)
))

# Execute grid search
ae_grid <- h2o.grid(
  algorithm = 'deeplearning',
  x = seq_along(features),
  training_frame = features,
  grid_id = 'autoencoder_grid',
  autoencoder = TRUE,
  activation = 'Tanh',
  hyper_params = hyper_grid,
  sparse = TRUE,
  ignore_const_cols = FALSE,
  seed = 123
)

# Print grid details
h2o.getGrid('autoencoder_grid', sort_by = 'mse', decreasing = FALSE)
```

Visualizing Reconstruction

```{r}
# Get sampled test images
index <- sample(1:nrow(mnist$test$images), 4)
sampled_digits <- mnist$test$images[index, ]
colnames(sampled_digits) <- paste0("V", seq_len(ncol(sampled_digits)))

# Predict reconstructed pixel values
best_model_id <- ae_grid@model_ids[[1]]
best_model <- h2o.getModel(best_model_id)
reconstructed_digits <- predict(best_model, as.h2o(sampled_digits))
names(reconstructed_digits) <- paste0("V", seq_len(ncol(reconstructed_digits)))

combine <- rbind(sampled_digits, as.matrix(reconstructed_digits))

# Plot original versus reconstructed
par(mfrow = c(1, 3), mar=c(1, 1, 1, 1))
layout(matrix(seq_len(nrow(combine)), 4, 2, byrow = FALSE))
for(i in seq_len(nrow(combine))) {
  image(matrix(combine[i, ], 28, 28)[, 28:1], xaxt="n", yaxt="n")
}
```

```{r}
ae100_codings <- h2o.deepfeatures(best_model, features, layer = 1)
ae100_codings %>% 
    as.data.frame() %>% 
    tidyr::gather() %>%
    summarize(average_activation = mean(value))
```

```{r}
# Hyperparameter search grid
hyper_grid <- list(sparsity_beta = c(0.01, 0.05, 0.1, 0.2))

# Execute grid search
ae_sparsity_grid <- h2o.grid(
  algorithm = 'deeplearning',
  x = seq_along(features),
  training_frame = features,
  grid_id = 'sparsity_grid',
  autoencoder = TRUE,
  hidden = 100,
  activation = 'Tanh',
  hyper_params = hyper_grid,
  sparse = TRUE,
  average_activation = -0.1,
  ignore_const_cols = FALSE,
  seed = 123
)

# Print grid details
h2o.getGrid('sparsity_grid', sort_by = 'mse', decreasing = FALSE)
```

```{r}
# Extract reconstruction errors
(reconstruction_errors <- h2o.anomaly(best_model, features))

# Plot distribution
reconstruction_errors <- as.data.frame(reconstruction_errors)

ggplot(reconstruction_errors, aes(Reconstruction.MSE)) +
  geom_histogram(aes(fill = ..count..), bins = 30)

@@
